### 2024-06-18 README.Rmd
More work

As research by @nezhurina_alice_2024 demonstrates, relatively simple natural language puzzles that are easily solved by humans, often result in confidently expressed nonsense when used to prompt an AI. Wrong answers, furthermore, are frequently backed up by confabulations, that is, reasoning-like or plausible sounding explanations that are illogical. If anything, mathematics should be easy: after all, computers have been doing mathematics since at least the [Antikythera mechanism](https://en.wikipedia.org/wiki/Antikythera_mechanism) (ca. 178 BC), which preceded other mechanical machines by centuries--including Babbage's [Analytical Engine](https://en.wikipedia.org/wiki/Analytical_engine), the machine that [Ada Lovelace](https://en.wikipedia.org/wiki/Ada_Lovelace) used to code an algorithm to write Bernoulli numbers. Electronic computers have been doing mathematics reliably for decades now. Computers can do mathematics well because they are implemented using a crisp, unambiguous language. Natural language is full of nuances that are not easily captured using the medium of text alone. The stochastic connections between words that LLMs produce are simply probabilistic word salads that only sound plausible thanks to the billions of data points used to train them and the billions of parameters that the models use. But if the objective of qualitative research is to identify interesting aspects of the human experience, it is certainly not the case that they will be found in a regression to the mean.

@@

But AIs must do something for someone, otherwise why would they be everywhere? What do AIs do for the likes of OpenAI, Microsoft, and Google? So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as the resources needed to have a University Committee to advise on AI^[Some observers have even wondered whether the AI fever is not actually becoming detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].]. Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].

@@

Alas, the industry has not necessarily made money for investors yet. For example, @berber_peter_2024 reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue. @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google. At a guess, the industry is running at a loss: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer. And a large number of users are not even paying subscribers: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023]. The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024]. The fact that the industry is not profitable has not prevented start-ups from pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted to AI from cryptocurrency, after cryptocurrency went sour with the frauds at FTX.].

Venture capital is not known for its charitable aims, though, and at some point investors will want their money back with a heap of profit on top [@sriram_ai_2023]. Who will pay to make investors whole? And how much will those who end up paying need to pay? Or, in the worst case, _have_ to pay, if their skills are inexorable tied to this one technology? Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020]. From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, instead of being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].

@@

### Extractive

In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )

@@

### Exploitative

ChatGPT 4.o was announced on May 13, 2024, with a flashy display of the model's new capabilities, which included the ability to respond vocally to prompts, and to identify images via a mobile device's camera. 4.o's voice, in particular, caused some titillation in the pants of even its creators [@roose_is_2024]. The voice was modeled after the digital personal assistant of the movie _Her_, played by Scarlett Johannson. This film (to approximately no one's surprise) is, by his own account, one of OpenAI's Altman's favorites. The background story of 4.o's voice is somewhat sordid: Altman reportedly tried to hire Ms. Johannson to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when Ms. Johannson declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways. This voice was retired (probably to the profound sadness of many an AI engineer) after Ms. Johansson threatened legal action^[Altman's denials that OpenAI had not used Ms. Johansson's voice sounded hollow, after he [posted](https://x.com/sama/status/1790075827666796666) on social media the word "her" at the time the new version of ChatGPT was announced.] [@allyn_scarlett_2024].

Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI. Most people in the world lack that power.

The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted. But AI companies do not spare the rich either.

@@

Think about linguistic hegemony, and how the models are only as good (or as bad) as their inputs.

@@

@@

## Loose thoughts

@@

Crypto-currency to AI: https://en.wikipedia.org/wiki/Cognition_AI

>>>>

As research by @nezhurina_alice_2024 demonstrates, relatively simple natural language puzzles that are easily solved by humans, often result in confidently expressed nonsense when used to prompt an AI. Wrong answers, furthermore, are frequently backed up by confabulations, that is, reasoning-like or plausible sounding explanations that are illogical. If anything, mathematics should be easy: after all, computers have been doing mathematics since at least the [Antikythera mechanism](https://en.wikipedia.org/wiki/Antikythera_mechanism) (ca. 178 BC), which preceded other mechanical machines by centuries--including Babbage's [Analytical Engine](https://en.wikipedia.org/wiki/Analytical_engine), the machine that [Ada Lovelace](https://en.wikipedia.org/wiki/Ada_Lovelace) used to code an algorithm to write Bernoulli numbers. Electronic computers have been doing mathematics reliably for decades now. Computers can do mathematics well because they are implemented using a crisp, unambiguous language. Natural language is full of nuances that are not easily captured using the medium of text alone. The stochastic connections between words that LLMs produce are simply probabilistic word salads^[@xu_hallucination_2024 use a formal approach to show that LLMs are inherently incapable of mapping outputs to a ground truth, and thus their results will only ever resemble a ground truth purely by chance. Recent reporting strongly hints that industry leaders like  Sundar Pichai and Tim Cook understand this well [@patel_google_2024; @tyrangiel_opinion_2024]] that only sound plausible thanks to the billions of data points used to train them and the billions of parameters that the models use. But if the objective of qualitative research is to identify interesting aspects of the human experience, it is certainly not the case that they will be found in a regression to the mean.

@@

But AIs must do something for someone, otherwise why would they be everywhere? What do AIs do for the likes of OpenAI, Microsoft, and Google? So far, the breakneck race to develop AI has soaked immense amounts of venture capital and other speculative money, in addition to incalculable resources by many other entities, such as those spent by universities to support an AI-dedicated committee^[Some observers have even wondered whether the AI fever is not actually becoming detrimental to innovation [@siegel_ai_2023; @vinsel_dont_2023].]. Between 2012 and 2020, venture capital investments in AI grew twenty five-fold, from \$3 billion (USD) to \$75 billion (USD)^[For perspective, Ontario's Ministry of Colleges and Universities, whose mission includes supporting "research and innovation to help the province compete and thrive in the global economy", spent $163,482,281 (CAD) on its research program in 2021-2022 [@ontario_ministry_of_colleges_and_universities_expenditure_2023], a mere blip (~0.0017\%) that barely registers compared to venture capital investments in AI in 2020 alone.], much of it poured into autonomous vehicles and related mobility tech, followed by healthcare, biotechnologies and drugs, with business processes and support services in third place [@tricot_venture_2021].

@@

Alas, the industry has not necessarily made money for investors yet.

@berber_peter_2024, for instance, reports that the AI industry spent \$50 billion (USD) in 2023 on chips alone, while bringing in only $3 billion in revenue. @patel_inference_2023 estimated that incorporating ChatGPT-like LLMs into search would lead to a loss of \$30 billion (USD) in profit for Google. At a guess, the industry is running at a loss in almost every respect: in the same report, @patel_inference_2023 calculated the cost of running a query in ChatGPT at about \$0.36--so it would take just about 57 queries per user with a "Plus" subscription ([\$20/month](https://openai.com/chatgpt/pricing/)) for OpenAI to begin losing money on that one customer. And a large number of users are not even paying subscribers: based on revenue reports [see @perez_chatgpts_2024] there were at most 210,000 subscribers to the "Plus" plan in 2024, out of a user base estimated at tens of millions [@tong_exclusive_2023]. The profitability of AI would look even worse if not for the fact that the cost of labor is suppressed by outsourcing the most intensive, disturbing, degrading, and damaging aspects of training AIs to the Global South [@nix_analysis_2024]. The fact that the industry is not profitable has not prevented start-ups from hurriedly pivoting to GenAI, in a shift  of focus to this much hyped technology from whatever they were doing in the first place^[As if any other signs  were needed of the scammy environment in which GenAI tech thrives, @berber_peter_2024 also reports how Cognition, a startup backed by Peter Thiel, pivoted from cryptocurrencyto AI, coincidentally as cryptocurrency was going sour after the frauds at FTX were revealed. Barely disguised grift has grown abundant in the vicinity of AI at almost every level [@cook_5_2024].].

Venture capital is not known for its charitable aims, though, and at some point investors will want their money back with a heap of profit on top [@sriram_ai_2023]. Who will pay to make investors whole? And how much those who end up paying will need to pay? Or, in the worst case, _have_ to pay, if their skills become inexorably tied to this one technology? Deskilling, both of individuals and of organizations, appears to be one of the few pathways to AI profitability^[And, with profitability becoming a more distant possibility, there is always the option for large investors to cosmically inflate the hype [@salmon_what_2024] to convince bigger suckers to give them money [@reuters_openais_2024] and/or to fleece retail investors [@temkin_vcs_2024].]: once people have failed to learn a skill, and therefore become dependent on the tool, it becomes possible for the maker of the tool to turn into a digital rentier [@sadowski_internet_2020]. From this perspective, what is on offer by the AI industry is SaaS--Skill as a Service^[Universities, rather than being innovators, have increasingly become renters in the land of the platform economy [@komljenovic_rise_2021].].

@@

### AI is an eminently extractive and exploitative industry

ChatGPT 4.o was announced on May 13, 2024, with fanfare and a flashy display of the model's new capabilities, which included the ability to respond vocally to prompts, and to identify images via a mobile device's camera. 4.o's voice, in particular, caused some titillation in the nether parts of even its creators [@roose_is_2024]. The voice was modeled after the digital personal assistant of the movie _Her_, played by Scarlett Johannson. This film (to approximately no one's surprise) is, by his own account, one of OpenAI's Altman's favorites. The background story of 4.o's voice is somewhat sordid: Altman reportedly tried to hire Ms. Johannson to be the voice of ChatGPT 4.o [@kastrenakes_scarlett_2024], but when Ms. Johannson declined, OpenAI went and used a mimicry of [her voice](@as-an-aside) anyways. This voice was retired (probably to the profound sadness of many an AI engineer) after Ms. Johansson threatened legal action^[Altman's denials that OpenAI had not used Ms. Johansson's voice sounded hollow, after he [posted](https://x.com/sama/status/1790075827666796666) on social media the word "her" at the time the new version of ChatGPT was announced.] [@allyn_scarlett_2024].

Alas, very few have the social standing of Scarlett Johansson to challenge OpenAI. Most people in the world lack that power. In relation to the points above (GenAIs require vast amounts of _human-generated_ data to ensure that there is sufficient variability for the model to train) - (Models trained )

@@

The ways in which GenAI is exploitative of people all over the world, but particularly in the Global South were highlighted.

@@

@@

@morrone_signals_2024 Carefully manufactured and well funded hype.

@@

@@

AI industry and the military: @haskins_im_2024
